---
title: "Reproducible Science and Figures Assignment"
format: html
embed-resources: true
editor: visual
---

### Data preparation:

```{r, message=FALSE}
library(tidyverse)
library(palmerpenguins)
library(janitor)
library(here)
library(ggplot2)
library(rstatix)
library(ggpubr)
library(ragg)
library(svglite)

source(here("functions", "cleaning.R"))
```

Loading the data:

```{r}
write.csv(penguins_raw, here("data", "penguins_raw.csv"))
```

Cleaning the data:

```{r}
penguins_clean <- penguins_raw %>%
  clean_column_names() %>%
  remove_columns(c("comments", "delta")) %>%
  shorten_species() %>%
  remove_empty_columns_rows()
```

Saving the cleaned data:

```{r}
write_csv(penguins_clean, here("data", "penguins_clean.csv"))
```

### Question 1: Data Visualisation for Science Communication

#### 1.1 Misleading figure:

```{r, message=FALSE}
body_mass_vs_flipper_length <- penguins_clean %>% 
  select(species, body_mass_g, flipper_length_mm) %>%
  remove_NA()

bm_vs_fl_scatterplot <- ggplot(data = body_mass_vs_flipper_length,
       aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(colour = species, fill = species), alpha = 0.25) +
  geom_smooth(aes(colour = species), method = "lm", se = FALSE) + # adding lines of best fit
  scale_x_continuous(limits = (c(2000,7000))) + # manually changing x axis scale
  scale_y_continuous(limits = (c(100,300))) + # manually changing y axis scale
  theme(legend.title = element_blank()) # removal of legend title

bm_vs_fl_scatterplot
```

1.1.1 Saving vectored figure

```{r, results="hide, warning=FALSE"}
ggsave(here("figures", "bm_vs_fl_vector.svg"),
       plot = bm_vs_fl_scatterplot,
       device = svglite,
       width = 20,
       height = 20,
       scaling = 2)

```

#### 1.2 Write about how your design choices mislead the reader about the underlying data (100-300 words)

All the data on this graph is correct, however, the design choices mean that the data is presented extremely poorly. The axis labels are not in a readable format as they have been left in the "snake case" format used in the data table. The axis ranges are unnecessarily large which masks the true distribution of the underlying data as the points are too close together. The legend has no title and it is unclear what the three categorical variables mean. Furthermore, the colours of three species makes it difficult to tell which data point belongs to which species, and makes it particularly difficult for colour blind people. Combining this with the high point transparency and the grey background it makes the distribution very difficult to discern. Finally, the lines or best fit for each species are not particularly clear and are not shown with any p-values or confidence intervals from the linear models.

### Question 2: Data Pipeline

#### 2.1 Introduction

The Palmer Penguins dataset contains information about 3 different penguin species (Adelie, Chinstrap, and Gentoo) which inhabit 3 islands (Biscoe, Dream, and Torgersen) of the Palmer Archipelago in Antarctica. The dataset contains information about individual penguins, including their species and the island they are found on. It also includes phenotypic information like sex, morphological measurements, and reproductive information. I will be exploring this dataset for results of statistical significance.

#### 2.2 Hypothesis

Before making any hypothesis, it is vital to explore the raw data to look for any patterns of particular biological interest:

Exploratory figure to investigate culmen length:

```{r}
culmen_length <- penguins_clean %>%
  select(species, culmen_length_mm) %>% # selecting required variables
  remove_NA()

# removing NA values from culmen length column
```

```{r}
species_colours <- c("Adelie" = "darkorange",
                     "Chinstrap" = "purple",
                     "Gentoo" = "cyan4")
# matching species with the colours of the dataset

exploring_culmen_length <- ggplot(
  data = penguins_clean %>% 
    select(species, culmen_length_mm, sex) %>% 
    remove_NA(),
  aes(x = species,
      y = culmen_length_mm)) +
  geom_violin(
    aes(colour = species, fill = species),
    draw_quantiles = c(0.25, 0.5, 0.75), # addition of medians and quartiles
    alpha = 0.2,
    trim = FALSE) +
  coord_flip() +
  geom_jitter(position = position_jitter(0.2, seed = 0), # set seed for reproducibility
              shape = 21,
              color = "black",
              fill = "white",
              size = 0.5) +
  scale_color_manual(values = species_colours) +
  scale_fill_manual(values = species_colours) +
  facet_grid(rows = vars(sex)) + # grouping plots by sex
  labs(x = "Species", y = "Culmen Length (mm)") +
  theme_bw()

exploring_culmen_length
```

::: {.callout-note icon="false"}
## Hypotheses

Looking at the data on **culmen length**, it would be interesting to determine whether **different species** have significant differences.

**Null hypothesis** - No difference in culmen length between species

**Alternative hypothesis** - There is a significant difference in culmen length between species
:::

#### 2.3 Statistical methods

2.3.1 Shapiro-Wilk test for normality

```{r}
shapiro_test <- shapiro.test(culmen_length$culmen_length_mm)

shapiro_test
```

The null hypothesis of the Shapiro-Wilk test is the data is drawn from a normally distributed population. As **p \< 0.05**, there is sufficient evidence to reject the null hypothesis and so culmen length is **not normally distributed**.

This can be visualised using a qqplot:

```{r}
qqnorm(culmen_length$culmen_length_mm)
qqline(culmen_length$culmen_length_mm, col = "red")
```

2.3.2 Kruskal-Wallis test

As culmen length is not normally distributed, a non-parametric method is required to test differences between species. The **Kruskal-Wallis** test determines whether there are significant differences between the medians of two or more groups. It requires a numerical response variable (culmen length) and a categorical explanatory variable (species).

```{r}
kruskal <- kruskal.test(data = culmen_length, culmen_length_mm ~ species)

kruskal
```

The null hypothesis of the Kruskal-Wallis test is that the medians are the same across all groups. As **p \< 0.05**, there is sufficient evidence to reject the null hypothesis and so at least one species has a different median culmen length.

2.3.3 Dunn test

Now it is has been established that at least one species has a significantly different median culmen length, pairwise interactions can be measured to determine how the culmen length individual species differ from each other. One such post-hoc test suitable for non-parametric data is the Dunn test.

```{r}
stats_dunn <- dunn_test(data = culmen_length, culmen_length_mm ~ species, p.adjust.method = "bonferroni")

stats_dunn_cleaned <- stats_dunn %>% 
  select(-n1, -n2, -statistic, -p)

stats_dunn_cleaned
```

Using the Dunn test, each pairwise comparison is its own hypothesis test. For each pair, the null hypothesis is that there is no difference between the medians. Multiple simultaneous hypothesis tests require a p-value adjustment to reduce Type I error - in this case the Bonferroni correction is used.

#### 2.4 Results & Discussion

```{r}
stats_dunn_annotation <- stats_dunn %>% 
    add_xy_position(x = "species") # annotating graph with results from Dunn test

culmen_length_violinplot <- ggplot(
  data = penguins_clean %>% drop_na(),
  aes(x = species,
      y = culmen_length_mm)) +
  geom_violin(
    aes(colour = species, fill = species),
    draw_quantiles = c(0.25, 0.5, 0.75),
    alpha = 0.2,
    trim = FALSE) +
  geom_dotplot(binaxis = "y", binwidth = 0.4, stackdir = "center", color = "black", fill = "white") +
  scale_color_manual(values = species_colours) +
  scale_fill_manual(values = species_colours) +
  stat_pvalue_manual(stats_dunn_annotation, hide.ns = FALSE) +
  theme_classic() +
  labs(x = "Species", y = "Culmen Length (mm)")
  

culmen_length_violinplot
```

This results figure clearly shows the pairwise comparisons and their significance from the Dunn test. Adelie penguins and Chinstrap penguins have a significantly different culmen lengths, as do Adelie compared with Gentoo. This is shown by the very small p-values represented by '\*\*\*\*'. There is also no significant difference between culmen lengths of Chinstrap compared to Gentoo penguins.

#### 2.5 Conclusion

We can conclude from the statistical analysis that culmen length differs among the three species, and specifically that there is difference between Adelie penguins compared with both Chinstrap and Gentoo. More research is required to determine the biological significance of this, as it may be due to scaling with body size or the different beak lengths could be a specific selective advantage.

### Question 3: Open Science

#### 3.1 GitHub link

<https://github.com/dusty-saxophone/Reproducible-Science-and-Figures-Assignment>

#### 3.2 Partner's GitHub link

#### 3.3 Reflection on partner's code

#### 3.4 Reflection on my code
