---
title: "Reproducible Science and Figures Assignment"
format: html
editor: visual
---

### Data preparation:

```{r}
library(tidyverse)
library(palmerpenguins)
library(janitor)
library(here)
library(ggplot2)

source(here("functions", "cleaning.R"))
```

#### Loading the data:

```{r}
write.csv(penguins_raw, here("data", "penguins_raw.csv"))
```

#### Cleaning the data:

```{r}
penguins_clean <- penguins_raw %>%
  clean_column_names() %>%
  remove_columns(c("comments", "delta")) %>%
  shorten_species() %>%
  remove_empty_columns_rows()
```

#### Saving the cleaned data:

```{r}
write_csv(penguins_clean, here("data", "penguins_clean.csv"))
```

### Question 1: Data Visualisation for Science Communication

#### a) Provide your figure here:

#### b) Write about how your design choices mislead the reader about the underlying data (100-300 words)

### 

### Question 2: Data Pipeline

#### Introduction

The Palmer Penguins dataset contains information about 3 different penguin species (Adelie, Chinstrap, and Gentoo) which inhabit 3 islands (Biscoe, Dream, and Torgersen) of the Palmer Archipelago in Antarctica. The dataset contains information about individual penguins, including species and the island they are found on. It also includes phenotypic information like sex, morphological measurements, and reproductive information.

#### Hypothesis

#### Statistical methods

#### Results & Discussion

#### Conclusion

### Question 3: Open Science
